### 2025-11-08 thesis/implementation/index.tex
Write about the benchmark system

Whilst the benchmarking infrastructure is not part of the core pipeline, they
help to develop and test hypotheses for extensions or design changes
confidently, compare performance across pipeline versions, models and
understand the benchmarking data.

>>>>

Whilst \textsc{Natural}'s benchmarking infrastructure is not part of the core
pipeline, it helps to develop and test hypotheses for extensions or design
changes confidently, compare performance across pipeline versions, models and
understand the benchmark performance.

\subsubsection{Execution-Based Evaluation System}

The \texttt{natural-benchmark} CLI evovles around the concept of executions
which are single benchmark runs against a specific benchmark dataset
(defaulting to \textsc{Spider}). An execution is the set of tests that have
been executed using a version of \textsc{Natural} against a benchmark. The
interface of \texttt{natural-benchmark} offers multiple options to create
executions and subsequentlty understand \textsc{Natural}'s performance:

\begin{enumerate}
    \item Running new benchmarks via: \\ \texttt{natural-benchmark new}
    \item Continuing a halted execution via: \\ \texttt{natural-benchmark continue --execution <id>}
    \item Comparing performance on previous (partial) executions via: \\ \texttt{natural-benchmark compare-to --previous <id>}
    \item Computing statistics on past executions via: \\ \texttt{natural-benchmark stats --execution <id>}
\end{enumerate}

While benchmarking \texttt{natural-benchmark} maintains a local SQLite database
maintaining a history of past performance, pipeline failures etc. for future
introspection as well as comparison of approaches.

\subsubsection{Cross-Dataset Validation}

In order to measure the performance of \textsc{Natural} against multiple
benchmarking datasets, the benchmarking and execution system in
\texttt{natural-benchmark} must be generalize across one dataset. The
benchmarking setup achieves this through a set of rust traits (eg, type
characteristics) to model benchmarking datasets.

The traits \texttt{Benchmark}, \texttt{BenchmarkDatabase} and
\texttt{BenchmarkTest} allow to abstract the file system layout of different
evaluation datasets. Thus when implementing benchmark support against
\textsc{Spider} the implementation absatracts the fact that \textsc{Spider} is
using SQLite for test databases, and uses a JSON file to store the questions.
The file system layout of \textsc{Spider} and \textsc{Bird} is largely similar.

This abstracted benchmarking system (see \ref{appendix:benchmark-traits})
allows integrating further benchmarks in the future (eg, \textsc{Spider2} which
relies on cloud databases like Snowflake) while maintaining the same ergonomics
and tooling across benchmark datasets (eg, recording and indexing all test
executions in a local database, recording pipeline logs etc).

\subsubsection{Metric Computation}

The benchmarking infrastructure computes the two metrics \textsc{Execution
Accuracy} and \textsc{Exact Match} found in prevalent benchmark leaderboards
(eg, \textsc{Spider} and \textsc{Bird}) through a Rust port of the Python
reference implementation found in \textsc{Spider}. \textsc{Spider} determines
the \textsc{Execution Accuracy} by comparing the result sets of two queries
based on possible row and column permutations as well as optional order
sensitivity. The \textsc{Exact Match} metric is determined by exact equivalence
of the ground truth query $\omega_{ground}$ and the candidate query $\omega$.

\subsubsection{Benchmarking Challenges}

Two primary issues emerged while developing the benchmarking infrastructure for
\textsc{Natural}:

While loading result sets from the provided test database, a faulty
code path failed to parse query result with types other than
\texttt{Text}. This casued empty result sets to be returned as query results
from both the ground truth query and the candidate query, resulting in a false
positive, skewing the \textsc{EA} metric significantly upwards.

Furthermore frequent pipeline failures of \textsc{Natural} made it hard to get
compute reliable performance metrics as context window constraints, out
of memory issues and CUDA issues caused a signficiant number of test
cases to fail, making subsequently computed metrics unreliable as only
a subset of the dataset was included in the results.
